---
title: "p8105_hw3_yl5219"
author: "Yuqing Liu"
date: "2023-10-14"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(ggplot2)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)


scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

### Problem 1

#### read in data 
```{r}
library(p8105.datasets)
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```

* a short description of the dataset:

  * The dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. 
  * Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. 
  * In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(user_id, order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.


* How many aisles are there, and which aisles are the most items ordered from?

In total, there are `r instacart |> select(aisle) |> distinct() |> count()`  aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

* Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

* Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

* Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```

### Problem 2

```{r}
library(p8105.datasets)
data("brfss_smart2010")

brfss_smart2010
```

First, do some data cleaning:

* format the data to use appropriate variable names by `janitor::clean_names()`;
* focus on the “Overall Health” topic by `filter(topic == "Overall Health")`
* include only responses from “Excellent” to “Poor” by ` filter(response %in% c("Excellent", "Very good", "Good", "Fair", "Poor"))`
* organize responses as a factor taking levels ordered from “Poor” to “Excellent” by `factor()` function

```{r}
brfss_smart2010 = brfss_smart2010 |> 
  as_tibble()|>
  janitor::clean_names()|>
  filter(topic == "Overall Health")|> 
  filter(response %in% c("Excellent", "Very good", "Good", "Fair", "Poor"))|>
  mutate(response = factor(response, levels = c("Excellent", "Very good", "Good", "Fair", "Poor")))

brfss_smart2010
```



Using this dataset, do or answer the following (commenting on the results of each):

* In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r}
brfss_smart2010|>
  filter(year == 2002)|>
  group_by(locationabbr)|>
  summarize(n = n_distinct(locationdesc))|>
  filter(n>=7)
```

In 2002, CT(Connecticut), FL(Florida), MA(Massachusetts), NC(North Carolina), NJ(New Jersey), PA(Pennsylvania) states were observed at 7 or more locations.

```{r}
brfss_smart2010|>
  filter(year == 2010)|>
  group_by(locationabbr)|>
  summarize(n = n_distinct(locationdesc))|>
  filter(n>=7)
```
In 2002, CA(California), CO(Colorado), FL(Florida), MA(Massachusetts),	MD(Maryland), NC(North Carolina), NE(Nevada), NJ(New Jersey), NY(New York), OH(Ohio) states were observed at 7 or more locations.

* Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r}
df = brfss_smart2010|>
  filter(response == "Excellent")|>
  group_by(year, locationabbr)|>
  summarize(data_value_mean=mean(data_value, na.rm = TRUE))

df
```
```{r}
df |> 
  ggplot(aes(x = year, y = data_value_mean, group = locationabbr, color = locationabbr)) +
  geom_line() + geom_point(size = 0.5)
```



* Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

```{r}
# Create a two-panel plot
brfss_smart2010 |>
  filter(locationabbr=="NY"& year %in% c(2006,2010))|>
  ggplot( aes(x = response, y = data_value)) +
  geom_boxplot() +
  facet_wrap(~year, ncol = 2) +
  labs(title = "Distribution of data_value by Response in New York State in 2006 and 2010",
       x = "Response",
       y = "data_value") +
  theme_minimal()
```


### Problem 3

* Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

  * load and tidy `nhanes_accel` dataset:
    * convert the dataset into long format by `pivot longer`
    * select only the value of `min` variable by `separate(min, into = c("prefix", "min"), sep = 3)` and `select(-prefix)`
  
```{r}
nhanes_accel = read.csv("./nhanes_accel.csv")|> 
  pivot_longer(
    min1:min1440,
    names_to = "min",
    values_to = "mims_value"
  )|>
  separate(min, into = c("prefix", "min"), sep = 3)|>
  select(-prefix)

nhanes_accel
```
  * load and tidy `nhanes_covar` dataset:
    * exclude participants less than 21 years of age by `filter(age >= 21)`
    * excluse those with missing demographic data by `drop_na()`
    * encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).
    
```{r}
nhanes_covar = read.csv("./nhanes_covar.csv", skip = 4)|>
  janitor::clean_names()|>
  as_tibble()|>
  filter(age >= 21)|>
  drop_na()|>
  mutate(education = case_match(education, 
                                1 ~ "less than high school",
                                2 ~ "high school equivalent",
                                3 ~ "more than high school"))|>
  mutate(sex = case_match(sex, 
                                1 ~ "male",
                                2 ~ "female"))

nhanes_covar
```

* Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.



* Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

* Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based 

